# The R Programming Environment {#rprog1}

```{r intro1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center')
library(tidyverse)
library(knitr)
library(bookdown)
library(scales)
library(kableExtra)
```

## Prerequisites

MECH 201

## Grading

### Grading for MECH 481A6

Course grades will be determined by the following five components:

```{r grading, echo=FALSE}
grading <- data.frame(a = c("Exams",
                      "Quizzes",
                      "Homework"),
                b = c(60, 25, 15))
knitr::kable(grading, 
             col.names = c("Assessment component", "Percent of grade"), 
             align = "c") %>%
             kable_styling(bootstrap_options = "striped", full_width = F)
```
## Course set-up

Please download and install the latest version of: R and RStudio (Desktop version,
Open Source edition). Both are free for anyone to download. 

Students will also need to download and install git software and create a GitHub account.

Here are useful links for this set-up: 

- R: https://cran.r-project.org 
- RStudio: https://www.rstudio.com/products/rstudio/#Desktop 
- Install git: https://git-scm.com/downloads (only ERHS 535 / 581A4)
- Sign-up for a GitHub account: https://github.com (only ERHS 535 / 581A4)

## Coursebook

This coursebook will serve as the only required textbook for this course. I regularly edit and add to this book, so content may change somewhat over the semester (particularly for later weeks. We typically cover about a chapter of the book every 1-2 weeks of the course.

This coursebook includes: 

- Links to the slides presented in class for each topic
- In-course exercises, typically including links to the data used in the exercise
- Homework assignments
- Appendix of [reference distributions](#dist)
- A list of vocabulary and concepts that should be mastered for each quiz

If you find any typos or bugs, or if you have any suggestions for how the book
can be improved, feel free to post it on the book's [GitHub Issues
page](https://github.com/SmogDr/edar_coursebook/issues).

This book was developed using Yihui Xie's [bookdown](https://bookdown.org) framework. The book is built using code that combines R code, data, and text to create a book for which R code and examples can be re-executed every time the book is re-built, which helps identify bugs and broken code examples quickly. The online book is hosted using GitHub's free [GitHub Pages](https://pages.github.com). All material for this book is
available and can be explored at [the book's GitHub
repository](https://github.com/SmogDr/edar_coursebook).

### Other helpful books (not required)

The best book to supplement the coursebook and lectures for this course is [R
for Data Science](http://r4ds.had.co.nz), by Garrett Grolemund and Hadley
Wickham. The entire book is freely available online through the same format at
the coursebook. You can also purchase a paper version of the book (published by
O'Reilly) through Amazon, Barnes & Noble, etc., for around $40. This book is an
excellent and up-to-date reference by some of the best R programmers in the
world.

There are a number of other useful books available on general R programming, including:

- [R for Dummies](https://colostate-primo.hosted.exlibrisgroup.com/primo-explore/fulldisplay?docid=01COLSU_ALMA51267598310003361&context=L&vid=01COLSU&lang=en_US&search_scope=Everything&adaptor=Local%20Search%20Engine&tab=default_tab&query=any,contains,r%20for%20dummies&sortby=rank&offset=0)
- [R Cookbook](https://colostate-primo.hosted.exlibrisgroup.com/primo-explore/fulldisplay?docid=01COLSU_ALMA21203304500003361&context=L&vid=01COLSU&lang=en_US&search_scope=Everything&adaptor=Local%20Search%20Engine&tab=default_tab&query=any,contains,r%20cookbook&sortby=rank&offset=0)
- [R Graphics Cookbook](http://www.amazon.com/R-Graphics-Cookbook-Winston-Chang/dp/1449316956/ref=sr_1_1?ie=UTF8&qid=1440997472&sr=8-1&keywords=r+graphics+cookbook)
- [Roger Peng's Leanpub books](https://leanpub.com/u/rdpeng)
- Various books on [bookdown.org](www.bookdown.org)

The R programming language is used extensively within certain fields, including
statistics and bioinformatics. If you are using R for a specific type of
analysis, you will be able to find many books with advice on using R for both
general and specific statistical analysis, including many available in print or
online through the CSU library.

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].
