# Strings, Dates, and Tidying {#rprog3}

```{r rprog2-1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center')
library(tidyverse)
library(lubridate)
library(knitr)
library(kableExtra)
library(bookdown)
```

```{r load_daily_show, echo = FALSE, message = FALSE}
daily_show <- read_csv(file = "data/daily_show_guests.csv", skip = 4)
daily_show <- rename(.data = daily_show,
                     year = YEAR,
                     job = GoogleKnowlege_Occupation, 
                     date = Show, 
                     category = Group,
                     guest_name = Raw_Guest_List)
```

## Objectives

This Chapter is designed around the following learning objectives. Upon completing this Chapter, you should be able to:  

- Describe how R stores the POSIXct date and time object internally
- Convert a column to a date format using `lubridate` functions
- Extract information from a date object (e.g., month, year, day of week) using `lubridate` functions
- Search, organize, and visualize data that are linked to date objects
- Manipulate character strings using the `stringr` and `tidyr` packages of functions
- Parse strings using regular expressions (regex)
- Define the primary characteristics of *"Tidy Data"*
- Apply functions from the `dplyr` and `tidyr` packages to make dataframes "tidy"

## Strings {#strings}
Strings are class of character vectors

### String detect & match

### Regular Expressions

### String split & separate

### String extract & remove

## Dates and Date-times

To begin, we discuss how the core R code `{base}` handles dates and times, since there is a ton of code out there that utilizes these older functions. We will then quickly transition to the `lubridate` family of functions (part of the Tidyverse) because of their versatility and ease-of-use.  

### Dates and Times in base R

Dates and times in base R all proceed from an *"epoch"* or *time origin*.  In R, the *epoch* or "dawn of time" occurred at midnight on January 1^st^, 1970.  For the sake of the R programming world, the concept of time started at that precise moment and has moved forward ever since.  To note: R can handle date-times before 1/1/1970; it just treats them as negative values!  

To see a date-time object, you can tell R to give you the current "System Time" by calling the `Sys.time()` function.

``` {r sys-time}
Sys.time()
```
As you can see, we got back the date, time, and current timezone used by my computer.  If you want to see how this time is stored in R internally, you can use `unclass()`, which returns an object value with its class attributes removed.  When we wrap `unclass()` around `Sys.time()`, we will see the number of seconds that have occurred between the epoch of 1/1/1970 and right now:

``` {r unclass-time}
unclass(Sys.time())
```

That's a lot of seconds.  How many years is that?  
Just divide that number by [60s/min $\cdot$ 60min/hr $\cdot$ 24hr/d $\cdot$ 365d/yr] ~ `r unclass(Sys.time())/60/60/24/365` years.  
This calculation ignores leap years but you get the point...

### Date-time formats
Note that the `Sys.time()` function provided the date in a ***"year-month-day"*** format and the time in an ***"hour-minute-second"*** format: `r Sys.time()`

Not everyone uses this exact ordering when they record dates and times, which is one of the reasons working with dates and times can be tricky.  You probably have little difficulty recognizing the following date-time objects as equivalent but not-so-much for some computer programs:

``` {r date-times-1, echo = FALSE}
date_time <- tibble(examples = c("12/1/99 8:46 PM",
                                 "1-Dec-1999 20: UTC",
                                 "December 1st, 1999, 20:46:00"))

knitr::kable(date_time, col.names = NULL, 
             caption = "Date-time objects come in different forms") %>%
  kable_styling(full_width = F,  bootstrap_options = "striped")

```

``` {block, type="rmdnote"}
You will often see time referenced with a **"UTC"**, which stands for *"Universal Time, Coordinated"*.  UTC is preferred by programmers  because it doesn't have a timezone and it doesn't follow *Daylight Savings Time* conventions (DST is the bane of many coders).

In practice, UTC is the same time as GMT (Greenwich Mean Time, pronounced "gren-itch") but with an important distinction.  GMT is one of the many [time-zones](https://wikipedia.org/wiki/List_of_tz_database_time_zones) across the planet, whereas, UTC has no time zone.
```

### Date-time classes in R

R has several classes of date-time objects, none of which are easy to remember:  

1. **`POSIXct`** - stored as the time, in seconds, between the `epoch` of 1970-01-01 00:00:00 UTC and the date-time object in question.  
    * the 'ct' stands for *"continuous time"* to represent "continuous seconds from origin";
    * A `POSIXct` object is a single numeric vector (and so provides for efficient computing).
2. **`POSIXlt`** - stored as a list of date-time objects.  
    * the 'lt' stands for *"list time"*.
    * A `POSIXlt` list contains the following elements:
      * *sec* as 0–61 seconds
      * *min* as 0–59 minutes
      * *hour* as 0–23 hours
      * *mday* as 1–31 day of the month
      * *mon* as 0–11 months after the first of the year
      * *year* as Years since 1900
      * *wday* as 0–6 day of the week, starting on Sunday
      * *yday* as 0–365 days of the year.
      * *isdst* as a flag for Daylight savings time.  Positive if in force, zero if not, negative if unknown.
3. **`POSIXt`** - this is a virtual class.  `POSIXt` (without the "l") is an internal way for R to convert between `POSIXct` and `POSIXlt` date-time objects.  
    * Think of the `POSIXt` as a way for R to perform operations/conversions between a `POSIXct` and `POSIXlt` object without throwing an error your way.

As a reminder, here are some of the most common **vector classes** in R:

``` {r vect-classes, echo = FALSE, warning = FALSE}

vect_classes <- tibble(classes = c("`character`",
                                   "`numeric`",
                                   "`factor`",
                                   "`Date`",
                                   "`logical`",
                                   "`date-time`"),
                       examples = c("\"Chemistry\", \"Physics\", \"Mathematics\"",
                                    "10, 20, 30, 40",
                                    "Male [underlying number: 1], Female [2]",
                                    "\"2010-01-01\" [underlying number: 14,610]",
                                    "TRUE, FALSE",
                                    "\"2020-06-23 11:05:20 MDT\""))
knitr::kable(vect_classes, col.names = c("Class", "Example")) %>%
  kable_styling(full_width = F,  bootstrap_options = "basic")
```

To discover the class of a vector (including a column in a dataframe -- remember
each column can be thought of as a vector), you can use `class()`:

```{r date-1, warning = FALSE}
class(Sys.time())
```

Both the `POSIXct` and `POSIXlt` class of objects return the same value to the user; the difference is really in how these classes store date-time objects internally.  To examine them, you can coerce `Sys.time()` into each of the two classes using `as.POSIXct` and `as.POSIXlt` functions and then examine their attributes.

``` {r ct_attr}
time_now_ct <- as.POSIXct(Sys.time())
unclass(time_now_ct)
```

``` {r lt_attr}
time_now_lt <- as.POSIXlt(Sys.time())
str(unclass(time_now_lt)) # the `str()` function makes the output more compact
```
It's easy to see why the `POSIXct` object class is more computationally efficient; but it's also nice to see all the date-time information packed into the `POSIXlt`.  This is why R keeps a key to unlock both using `POSIXt`.  As my father used to say: clear as mud?

### Reading and classifying date-times
Oftentimes, when data is read into R, there are column elements that contain date and time information.  These dates and times are often interpreted by R as *character* vectors, which means they have lost their relational attributes (i.e., you cannot subtract "Monday 08:00" from "Wednesday 12:00" and get "2 days 4 hours"). If we want to analyze dates and times in a relational way, we need to instruct R to recognize these as date-time objects (i.e., as either the `POSIXct` or `POSIXlt` class). Thus, to convert a character vector into date or date-time object requires a change of that vector's ***class***.

Date-time elements can be tricky to work with for a few reasons:  

1. Different programs store and handle dates and times in different ways  
2. The existence of time zones means that date-time values can change with location
3. Date-time strings can be separated with spaces, colons, commas, slashes, dashes, or a mix of all those together (see Table \@ref(tab:date-times-1))

The `{base}` R function to convert between `character` classes and `date-time` classes is the function `strptime()`, which is short for *"**str**ing **p**arse into date-**time**"*. I mention this function not because I encourage you to use it but because I want you to be able to recognize it.  The function has over 39 conversion specifications that it can take as arguments.  That is to say, this function not simple to master.  If you are a glutton for punishment, I invite you to read the R Documentation `?strptime`.

In **summary** here are a few `{base}` R functions on date-time object that are worth knowing:

``` {r base-r-times, echo = FALSE}

base_r_times <- tibble(functions = c("Sys.Date()", 
                                     "Sys.time()",
                                     "Sys.timezone()",
                                     "as.POSIXct()",
                                     "as.POSIXlt()",
                                     "strptime()"),
                       returned = c("Current system date",
                                    "Current system date-Time",
                                    "Current system timezone",
                                    "date-time object of class POSIXct",
                                    "date-time object of class POSIXlt",
                                    "date-time object of class POSIXlt"),
                       examples = c("\"2020-06-23\"",
                                    "\"2020-06-23 11:05:20 MDT\"",
                                    "\"America/Denver\"",
                                    "\"2020-06-23 11:05:20 MDT\"",
                                    "\"2020-06-23 11:05:20 MDT\"",
                                    "\"2020-06-23 11:05:20 MDT\"")
)

knitr::kable(base_r_times, col.names = c("{base} R Function",
                                         "Value Returned",
                                         "Example"), 
             caption = "Basic Date-time functions") %>%
  kable_styling(full_width = F,  bootstrap_options = "striped")
```

## `Lubridate` {#lubridate}
The `lubridate` package was developed specifically to make life easier when working with date-time objects. 

### `Lubridate` Parsing Functions
One of best aspects of `lubridate` is its ability to parse date-time objects with simplicity and ease; the `lubridate` parsing functions are designed as "named-to-order". Let me explain:  

> <span style="color: purple;"> **Parse**: ***to break apart and analyze the individual components*** (of something, like a character string) </span>

* If a character vector is written in "**y**ear-**m**onth-**d**ay" format (i.e., `"2020-Dec-18"`), then the `lubridate` function to convert that vector is `ymd()`.

* If a character vector is written in "**d**ay-**m**onth-**y**ear" format (i.e., `"18-Dec-2020"`), then the `lubridate` function to convert that vector is `dmy()`.  Try it out:

``` {r lubridate-1}
#create a character vector
date_old <- "2020-Dec-18"

#prove to yourself it's a character class
class(date_old)

#convert it to a `Date` class with `ymd()`
date_new <- ymd(date_old)

#prove to yourself it worked
class(date_new)
```

That little conversion exercise may not have blown you away, but watch what happens when I feed the following set of wacky character vectors into that same `lubridate` parsing function, `ymd()`:

``` {r lubridate-2}

messy_dates <- c("2020------Dec the 12",
                 "20.-.12.-.12",
                 "2020aaa12aaa12",
                 "20,12,12",
                 "2020x12-12",
                 "2020   ....    12        ......     12",
                 "'20.December-12")

ymd(messy_dates)
```

That's right.  The `ymd()` parsing function figured them all out correctly with almost no effort on your part.  But wait, there's more!  The `lubridate` package contains parsing functions for almost any order you can dream up.  

``` {r lubridate-table1, echo = FALSE}

lubridate_parsers <- tibble(functions = c("`ymd()`",
                                           "`mdy()`",
                                           "`dmy()`"),
                            formats = c("year-month-day",
                                        "month-day-year",
                                        "day-month-year")
)


knitr::kable(lubridate_parsers, col.names = c("Parsing Function", "Format to Convert")) %>%
  kable_styling(full_width = F,  bootstrap_options = "basic")
```

And if you need to parse a time component, simply add a combination of `_hms` to the function call to parse time in "hours-minutes-seconds" format.  Some additional examples of how you would parse time that followed from a `ymd` format:

``` {r lubridate-table2, echo = FALSE}

lubridate_parsers2 <- tibble(functions = c("`ymd_h()`",
                                           "`mdy_hm()`",
                                           "`dmy_hms()`"),
                            formats = c("year-month-day_hours",
                                        "year-month-day_hours-minutes",
                                        "year-month-day_hours-minutes-seconds")
)

knitr::kable(lubridate_parsers2, col.names = c("Parsing Function", "Format to Convert")) %>%
  kable_styling(full_width = F,  bootstrap_options = "basic")
```
The beauty of the `lubridate` parsers is that they do the hard work of cleaning up the character vector, regardless of separators or delimiters within each string, and return either a `Date` or `Date-time` object class.

### Date-time manipulation with `Lubridate`

To convert the `date` column in the `daily_show` data into a Date
class, then, you can run:

```{r}
library(package = "lubridate")

class(x = daily_show$date) # Check the class of the 'date' column before mutating it

daily_show <- mutate(.data = daily_show,
                     date = mdy(date))
head(x = daily_show, n = 3)
class(x = daily_show$date) # Check the class of the 'date' column after mutating it
```

Once you have an object in the `Date` class, you can do things like plot by
date, calculate the range of dates, and calculate the total number of days the
dataset covers:

```{r eval = FALSE}
# report the min and max dates
range(daily_show$date)

# calculate the duration from first to last date
diff(x = range(daily_show$date)) 
```

We could have used these to transform the date in `daily_show`, using the following pipe chain: 

```{r message = FALSE}
daily_show <- read_csv(file = "data/daily_show_guests.csv",
                       skip = 4) %>%
  rename(job = GoogleKnowlege_Occupation, 
         date = Show,
         category = Group,
         guest_name = Raw_Guest_List) %>%
  select(-YEAR) %>%
  mutate(date = mdy(date)) %>%
  filter(category == "Science")
head(x = daily_show, n = 2)
```

The `lubridate` package also includes functions to pull out certain elements of a date, including: 

- `wday` return the day of the week pertaining to a Date object
- `mday` return the day of the month pertaining to a Date object
- `yday` return the day of the year pertaining to a Date object
- `month` return the month pertaining to a Date object
- `quarter` return the quarter of hte year pertaining to a Date object
- `year` return the year pertaining to a date object

For example, we could use `wday` to create a new column with the weekday of each show: 

```{r}
mutate(.data = daily_show,
       show_day = wday(x = date, label = TRUE)) %>%
  select(date, show_day, guest_name) %>%
  slice(1:5)
```

```{block, type = "rmdwarning"}
R functions tend to use the timezone of **YOUR** computer's operating system by
default, or UTC, or GMT. You need to be careful when working with dates and
times to either specify the time zone or convince yourself the default behavior
works for your application.
```


## Tidy Data



